---
title: "MCEM part2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, include=FALSE}
library(dmea)
```

From [this report]() we know that in the case of monte-carlo with 10 trees, $\lambda=0.8$, $\beta=0.0175$ and $\mu=0.1$ simulated trees, the estimations of MCEM are really good for $\lambda$ and $\beta$, but slyghtly biased for $\mu$

```{r}
0.1-0.08918 
```

we complain that if $\mu$ would not be biased, then we would have perfect MCEM performance. Our trys on finding the right re-weithing method to make $\mu$ unbiased have not been succeful yet. However, we can probe if our complains make senses or not. 

To do that, we just add the correct number to make it unbiased...


```{r correctmu1}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(8,0.175,0.9),n_trees=10,dummy=0.1-0.08918,n_it = 500,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

***

if we do it normally?

```{r normal EM}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(8,0.175,0.9),n_trees=10,dummy=0,n_it = 500,printpar = F)
```


the $E[l]$ value
```{r}
plot(em$El)
```

tha parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```


***

what if we try the adjusted one with other initial values?
```{r correctmu2}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(0.5,0.02,0.2),n_trees=10,dummy=0.1-0.08918,n_it = 500,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

***

normally, but with thoso others parameters?

```{r normal EM2}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(0.5,0.02,0.2),n_trees=10,dummy=0,n_it = 500,printpar = F)
```


the $E[l]$ value
```{r}
plot(em$El)
```

tha parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

nothing...

***

if we start with rates greater than 1?


what if we try the adjusted one with other initial values?
```{r correctmu3}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(1.1,0.02,1),n_trees=10,dummy=0.1-0.08918,n_it = 500,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

or

```{r}
sa <- stack(as.data.frame(em$Pars))
sa$x <- rep(seq_len(nrow(em$Pars)), ncol(em$Pars))
qplot(x, values, data = sa, group = ind, colour = ind, geom = "line")
```
***

(I will start doing less number of iterations)



now even greater rates

what if we try the adjusted one with other initial values?
```{r correctmu4}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(1.6,0.02,1.2),n_trees=10,dummy=0.1-0.08918,n_it = 200,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

or

```{r}
sa <- stack(as.data.frame(em$Pars))
sa$x <- rep(seq_len(nrow(em$Pars)), ncol(em$Pars))
qplot(x, values, data = sa, group = ind, colour = ind, geom = "line")
```

and

```{r}
summary(tail(em$Pars[,3],n = 100))
summary(tail(em$Pars[,1],n = 100))
```
***

if we (cheat less and)  do not take the perfect adjustment?

```{r correctmu5}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(1.6,0.02,1.2),n_trees=10,dummy=(0.1-0.08918)/2,n_it = 400,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

or

```{r}
sa <- stack(as.data.frame(em$Pars))
sa$x <- rep(seq_len(nrow(em$Pars)), ncol(em$Pars))
qplot(x, values, data = sa, group = ind, colour = ind, geom = "line")
```

and

```{r}
summary(tail(em$Pars[,3],n = 200))
summary(tail(em$Pars[,1],n = 200))
plot(tail(em$Pars[,3],n = 300))
```


it`s weird, it was suppose to go to zero, but it also stabilized the EM iterations. 

let`s see if for a minimal dummy value we still have this


if we (cheat less and)  do not take the perfect adjustment?

```{r correctmu6}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(1.6,0.02,1.2),n_trees=10,dummy=0.00001,n_it = 400,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

or

```{r}
sa <- stack(as.data.frame(em$Pars))
sa$x <- rep(seq_len(nrow(em$Pars)), ncol(em$Pars))
qplot(x, values, data = sa, group = ind, colour = ind, geom = "line")
```

and

```{r}
summary(tail(em$Pars[,3],n = 200))
summary(tail(em$Pars[,1],n = 200))
```

ok, not anymore..
***

let´s see for bigger numbers

```{r correctmu7}
st <- sim_phyl()
plot(st$newick,align.tip.label=T,root.edge=T)
st2 <- phylo2p(drop.fossil(st$newick))
em <- EM_phylo(wt=st2$t,init_par=c(1.6,0.02,1.2),n_trees=10,dummy=0.2, n_it = 400,printpar = F)
```


```{r}
plot(em$El)
```

the parameters
```{r}
plot(em$Pars[,1])
plot(em$Pars[,2])
plot(em$Pars[,3])
```

or

```{r}
sa <- stack(as.data.frame(em$Pars))
sa$x <- rep(seq_len(nrow(em$Pars)), ncol(em$Pars))
qplot(x, values, data = sa, group = ind, colour = ind, geom = "line")
```

and

```{r}
summary(tail(em$Pars[,3],n = 200))
summary(tail(em$Pars[,1],n = 200))
```
