---
title: ""
author: ""
date: "January 30, 2016"
output: beamer_presentation
---

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```

## Slide with Plot

```{r, echo=FALSE}
library(ape)
sl = paste(letters[1],letters,":0",sep="")
for (i in 2:26){
  ll = paste(letters[i],letters,":0",sep="")  
  sl = c(sl,ll)
}


phyl <- function(nT=100,theta=c(3,4),phi=c(1,2), draw=TRUE, model="loglinear"){
 # set.seed(1)
  traits = list()
  x=1
  tree = NULL
  a = rexp(1,4) # covariate for first specie
  A = a
  i <- 1
  Tm=0
  newick = paste(sl[1],";",sep="")
  newickm = list( substr(sl[1],1,2))
  E=0
    while (i<=nT){
      if (model == "loglinear"){
        lambda <- exp(theta[1]+theta[2]*A)     # Covariates model
        mu <- exp(phi[1]+phi[2]*A)             # Covariates model
      }
      if (model == "linear"){
        lambda <- theta[1]+theta[2]*A
        mu <- phi[1]+phi[2]*A
      }
      if (model == "logistic"){
        lambda <- theta[1]/(1+exp(-theta[2]*A))
        mu <- phi[1]/(1+exp(-phi[2]*A))
      }
      s = sum(lambda)+sum(mu)
      tm <- rexp(1,s)
      Tm[i] = tm
      prob <- c(lambda,mu)/s
      BD <- sample(2*x,1,prob=prob)  #speciation/extinction--specie
      BD
      if(BD>x){   #extinction
        a[i] = A[BD-x] 
        A <- A[-(BD-x)] # remove the value for the extinct specie
        specie = newickm[BD-x]
        ind = regexpr(specie,newick)[1] + 2
        newick = paste(substr(newick,1,ind),as.character(tm),substring(newick,ind+2),sep="")
        newickm[[BD-x]] <- NULL
        E[i] = 0
        x = x - 1
      } else {  # speciation
        a[i] = A[BD] 
        A <- c(A,A[BD]*rgamma(1,100,100))  #quizas seria mejor una normal... o algo simetrico que no pueda ser negativo
        specie = newickm[BD]
        ind = regexpr(specie,newick)[1] -1
        newick <- paste(substr(newick,1,ind),"(",sl[i],",",sl[677-i],"):",as.character(tm),substring(newick,ind+5),sep="")
        newickm[BD] <- substr(sl[i],1,2)
        newickm <-append(newickm,list(sl[677-i]))
        nspe = x + 1  #?
        x = x + 1
        E[i] = 1
      }
    if (x==0){ # In case we have all species extinct: restart
      x=1
      tree = NULL
      a = rexp(1,4) # covariate for first specie
      newick = paste(sl[1],";",sep="")
      rm(newickm)
      newickm = list( substr(sl[1],1,2))
      A=a
      i <- 1
    } else { # Otherwise, update values and go to next iteration (... and maybe plot)
      traits[[i]]<-A
      i<-i+1
    }
  
      }
#  newick <- gsub(":0)",":0.004)",newick)
#  newick <- gsub(":0,",":0.04,",newick)
  if(draw){
    ppp = read.tree(text=newick)
    plot(ppp)
  }
  treeD = list(t=Tm, cov = a, traits=traits, E=E, newick=newick)
}

fn1 <- function(theta, nT = 100) {
  parm = exp(theta[1] + theta[2]*a)*E+exp(theta[3] + theta[4]*a)*(1-E)
  sig = 0
  for (i in 1:nT){
    sig[i] = sum(exp(theta[1] + theta[2]*A[[i]])) + sum(exp(theta[3] + theta[4]*A[[i]]))
  }
  fn=sum(-sig*Tm+log(parm))
  -fn
}

fn2 <- function(theta, nT = 100) {
  parm = (theta[1] + theta[2]*a)*E+(theta[3] + theta[4]*a)*(1-E)
  sig = 0
  for (i in 1:nT){
    sig[i] = sum((theta[1] + theta[2]*A[[i]])) + sum((theta[3] + theta[4]*A[[i]]))
  }
  fn=sum(-sig*Tm+log(parm))
  -fn
}

fn3 <- function(theta, nT = 100) {
  parm = (theta[1]/(1+exp(-theta[2]*a)))*E+(theta[3]/(1+exp(-theta[4]*a)))*(1-E)
  sig = 0
  for (i in 1:nT){
    sig[i] = sum((theta[1]/(1+exp(-theta[2]*A[[i]])))) + sum((theta[3]/(1+exp(-theta[4]*A[[i]]))))
  }
  fn=sum(-sig*Tm+log(parm))
  -fn
}


```

```{r}
model = "loglinear"
s1 = phyl(nT=10, model = model)
```


## Tables

```{r, results='asis', xtable.margin = TRUE}
rm(P)
nP=100
P <- matrix(nrow=nP,ncol=4)
for (j in 1:nP){
  model = "loglinear"
  fn <- fn1
  s1 = phyl(nT=100, model = model, draw = FALSE)
  E = s1$E
  A = s1$traits
  a = s1$cov
  Tm = s1$t
  length(E[E==0])/length(E[E==1])
  op <- nlm(fn, theta <- c(1,1,1,1), hessian=TRUE)   # Initial values!! ?, nlm
  P[j,1]=op$estimate[1]
  P[j,2]=op$estimate[2]
  P[j,3]=op$estimate[3]
  P[j,4]=op$estimate[4]
}
p1 <- na.omit(P[,1])
p2 <- na.omit(P[,2])
p2t <- p2[abs(p2)<30]
p3 <- na.omit(P[,3])
p4 <- na.omit(P[,4])
p4t <- p4[abs(p4)<30]
S <- matrix(nrow=4,ncol=6)
colnames(S) <- c("n","real value","mean","median","min","max")
L=list(p1,p2,p3,p4)
l=c(3,4,1,2)
for (j in 1:4){
  p=L[[j]]
  S[j,1] = length(p)
  S[j,2] = l[j]
  S[j,3] = mean(p)
  S[j,4] = median(p)
  S[j,5] = min(p)
  S[j,6] = max(p)
}
library(xtable)
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(S, caption = "Model 1")
```