---
title: "What extant species can tell us about extinction?"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, include=FALSE, cache=FALSE}
library(dmea)
library(matrixStats)
```




```{r, echo=FALSE, include=FALSE}
mu = 0.1
pars = c(0.8,mu,40)
ltt = expectedLTT(pars)
qplot(ltt$t,ltt$Ex)
ltt = expectedLTT(pars,drop.extinct = TRUE)
qplot(ltt$t,ltt$Ex)
```

Because extinct species are rarelly included on phylogenetic trees, we are interested on invetigate the information that extant species contains about extinction rates. On the plot below we can see the expected Ltt plot, of extant-species only trees, for 3 different extinction rates


```{r, c1, echo=FALSE}
mu = 0.1
pars = c(0.8,mu,40)
Ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
mu = 0.2
pars = c(0.8,mu,40)
ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
Ltt = rbind(Ltt,ltt)
mu = 0.3
pars = c(0.8,mu,40)
ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
Ltt = rbind(Ltt,ltt)
Ltt$mu = as.factor(Ltt$mu)
  
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() +   geom_point(size=1)
```

we can see a clear difference on Ltt plots of extant species, smaller extinction rates tents to grow faster on the begining whereas higger extinction rate seems to have a slow grow on the begining. 

It seems also that is a matter of the first derivate, we can look at that also

```{r, echo=FALSE}
l1 = Ltt[Ltt$mu == 0.1,]
dif1 = diff(l1$Ex)
l1 = l1[-1,]
l1$Ex = dif1

l2 = Ltt[Ltt$mu == 0.2,]
dif2 = diff(l2$Ex)
l2 = l2[-1,]
l2$Ex = dif2

l3 = Ltt[Ltt$mu == 0.3,]
dif3 = diff(l3$Ex)
l3 = l3[-1,]
l3$Ex = dif3

Ltt = rbind(l1,l2,l3)
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() + geom_point()
```


Now we do it again, but with $K=80$ rather than $K=40$ in order to check some influence on the $K$ parameter

```{r, c2, echo=FALSE}
mu = 0.1
pars = c(0.8,mu,80)
Ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
mu = 0.2
pars = c(0.8,mu,80)
ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
Ltt = rbind(Ltt,ltt)
mu = 0.3
pars = c(0.8,mu,80)
ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
Ltt = rbind(Ltt,ltt)
Ltt$mu = as.factor(Ltt$mu)
  
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() +   geom_point()
```

It seems it is just a change on scale. Now, what about $\lambda$?

We set $=\lambda = 1.2$ rather than $\lambda=0.8$ and we see again the ltt plot

```{r, c3, echo=FALSE}
mu = 0.1
pars = c(1.2,mu,40)
Ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
mu = 0.2
pars = c(1.2,mu,40)
ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
Ltt = rbind(Ltt,ltt)
mu = 0.3
pars = c(1.2,mu,40)
ltt = data.frame(expectedLTT(pars,drop.extinct = TRUE),mu=mu)
Ltt = rbind(Ltt,ltt)
Ltt$mu = as.factor(Ltt$mu)
  
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() +   geom_point()
```
and we check the derivative again

```{r, echo=FALSE}
l1 = Ltt[Ltt$mu == 0.1,]
dif1 = diff(l1$Ex)
l1 = l1[-1,]
l1$Ex = dif1

l2 = Ltt[Ltt$mu == 0.2,]
dif2 = diff(l2$Ex)
l2 = l2[-1,]
l2$Ex = dif2

l3 = Ltt[Ltt$mu == 0.3,]
dif3 = diff(l3$Ex)
l3 = l3[-1,]
l3$Ex = dif3

Ltt = rbind(l1,l2,l3)
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() + geom_point()
```


# 2 parameter estimation (fixing $\mu$) 

```{r, ex2, echo=FALSE}
p = proc.time()
n_it = 1000
mu = 0.1
n_trees=10
MP = matrix(nrow=n_it, ncol=3)
RP = matrix(nrow=n_it, ncol=3)
for (i in 1:n_it){
  s = sim_phyl()
  p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
  RP[i,] = p
  wt = (s$newick.extant.p)$wt
  trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
  pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
  MP[i,] = c(pars[1],mu,pars[2])
}
par_est_vis(P=MP,par=1,PR=RP)
par_est_vis(P=MP,par=3,PR=RP)
print(proc.time()-p)
```

```{r, ex22, echo=FALSE}
p = proc.time()
n_it = 1000
mu = 0.1
n_trees=100
MP = matrix(nrow=n_it, ncol=3)
RP = matrix(nrow=n_it, ncol=3)
for (i in 1:n_it){
  s = sim_phyl()
  p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
  RP[i,] = p
  wt = (s$newick.extant.p)$wt
  trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
  pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
  MP[i,] = c(pars[1],mu,pars[2])
}
par_est_vis(P=MP,par=1,PR=RP)
par_est_vis(P=MP,par=3,PR=RP)
print(proc.time()-p)
```

Ok. The estimations are fine. 

Now let´s try for a grid of $\mu$

```{r, ex3}
mu0 = seq(0.04,0.3,by=0.01)
s = sim_phyl(seed=3)
p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
wt = (s$newick.extant.p)$wt
MP = matrix(nrow=length(mu0), ncol=3)
n_trees = 10
for(i in 1:length(mu0)){
  mu = mu0[i]
  trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
  pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
  MP[i,] = c(pars[1],mu,pars[2])
}
qplot(MP[,1],MP[,2])
```


```{r}
lm1 = lm(MP[,2] ~ MP[,1])
summary(lm1)
l1 = MP[,1]
m1 = MP[,2]
k1 = MP[,3]
MP1=MP
```

Does it help 100 trees (probably not)

```{r, ex32}
mu0 = seq(0.04,0.3,by=0.01)
s = sim_phyl(seed=3)
p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
wt = (s$newick.extant.p)$wt
MP = matrix(nrow=length(mu0), ncol=3)
n_trees = 100
for(i in 1:length(mu0)){
  mu = mu0[i]
  trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
  pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
  MP[i,] = c(pars[1],mu,pars[2])
}
qplot(MP[,1],MP[,2])
```

```{r}
lm2 = lm(MP[,2] ~ MP[,1])
summary(lm2)
l2 = MP[,1]
m2 = MP[,2]
k2 = MP[,3]
MP2=MP
```
Actually, the variance decreases. 

1000 trees?

```{r, ex33}
mu0 = seq(0.04,0.3,by=0.01)
s = sim_phyl(seed=3)
p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
wt = (s$newick.extant.p)$wt
MP = matrix(nrow=length(mu0), ncol=3)
n_trees = 1000
for(i in 1:length(mu0)){
  mu = mu0[i]
  trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
  pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
  MP[i,] = c(pars[1],mu,pars[2])
}
qplot(MP[,1],MP[,2])
```

```{r}
lm3 = lm(MP[,2] ~ MP[,1])
summary(lm3)
l3 = MP[,1]
m3 = MP[,2]
```

10000 trees?

```{r, ex34}
mu0 = seq(0.04,0.3,by=0.01)
s = sim_phyl(seed=3)
p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
wt = (s$newick.extant.p)$wt
MP = matrix(nrow=length(mu0), ncol=3)
n_trees = 10000
for(i in 1:length(mu0)){
  mu = mu0[i]
  trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
  pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
  MP[i,] = c(pars[1],mu,pars[2])
}
qplot(MP[,1],MP[,2])
```


```{r}
lm4 = lm(MP[,2] ~ MP[,1])
summary(lm4)
l4 = MP[,1]
m4 = MP[,2]
```


let`s look at them all together

```{r}
regressions = data.frame(mu = rep(m1,4), l = c(l1,l2,l3,l4), r = c(rep(x = 10,length(m1)),rep(x = 100,length(m1)),rep(x = 1000,length(m1)),rep(x = 10000,length(m1)) ) )
regressions$r = as.factor(regressions$r)
ggplot(data=regressions, aes(x=l,y=mu,colour=r)) + geom_point() + geom_smooth(method='lm')
```

Just for convenience: $\mu$ vs $\lambda$

```{r}
ggplot(data=regressions, aes(x=mu,y=l,colour=r)) + geom_point() + geom_smooth(method='lm')
```

get coefficients

```{r}
lm11 = lm(l1~m1)
lm22 = lm(l2~m2)
lm33 = lm(l3~m3)
lm44 = lm(l4~m4)
lm11$coef
lm22$coef
lm33$coef
lm44$coef
lm11$coef[1] + lm11$coef[2]*0.1
lm22$coef[1] + lm22$coef[2]*0.1
lm33$coef[1] + lm33$coef[2]*0.1
lm44$coef[1] + lm44$coef[2]*0.1
```

\newpage

# The Ltt plot

Now let`s try to minimize ltt, but first vizualize it

```{r, ex4}
time = proc.time()
grid = wt
Ltt = data.frame(t=cumsum(wt), Ex = (s$newick.extant.p)$n, mu=999)
for(i in 1:length(mu0)){
  mu = mu0[i]
  pars = c(l1[i],m1[i],k1[i]) 
  ltt = data.frame(t = cumsum(wt), Ex = expectedLTT(pars,drop.extinct = TRUE, grid=cumsum(wt))$Ex, mu=mu)
  Ltt = rbind(Ltt,ltt)
}
Ltt$mu = as.factor(Ltt$mu)
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() +   geom_point()
print(proc.time()-time)
```

```{r, ex48}
time = proc.time()
Ltt2 = data.frame(t=cumsum(wt), Ex = (s$newick.extant.p)$n, mu=999)
for(i in 1:length(mu0)){
  mu = mu0[i]
  pars = c(l1[i],m1[i],k1[i]) 
  expe = expectedLTT2(pars)
  ltt2 = data.frame(t = cumsum(expe$t), Ex = expe$Ex, mu=mu)
  Ltt2 = rbind(Ltt2,ltt2)
}
Ltt2$mu = as.factor(Ltt2$mu)
print(proc.time()-time)
```

```{r}
ggplot(data=Ltt2, aes(x=t, colour = mu)) + geom_step(aes(y=Ex)) #+  scale_color_brewer(name = "Method", palette = "Set1")
```


```{r, ex42}
ltt1 = Ltt[Ltt$mu == 999,]
diff_ltt = NaN
sq_diff = NaN
diff_mat = matrix(0,ncol=length(mu0),nrow=length(wt))
for(i in 1:length(mu0)){
 mu = mu0[i]
 ltt = Ltt[Ltt$mu == mu,]
 ltt$Ex = abs(ltt1$Ex-ltt$Ex)
 diff_mat[,i] = ltt$Ex
 diff_ltt[i] = sum(ltt$Ex)
 sq_diff[i] = sum((ltt1$Ex-ltt$Ex)^2)
 
}
n_diff_ltt = diff_ltt*wt
n_diff_ltt
diff_ltt_M = data.frame(mu = mu0, diff_ltt = diff_ltt)
choosed_mu = diff_ltt_M[diff_ltt_M$diff_ltt == min(diff_ltt_M$diff_ltt) ,]
choosed_mu$mu
b = matrix(0,ncol=length(mu0),nrow=length(wt))
for(i in 1:length(mu0)){
   b[i,] = diff_mat[i,]==rowMins(diff_mat)[i]
}
colSums(b)
MP[MP[,2] == choosed_mu$mu]
MP1[max(colSums(b)) == colSums(b)]
b[,max(colSums(b)) == colSums(b)]
```

Let´s use the second method

```{r, ltt2sim, cache=FALSE}
stat=NULL
for(i in 1:length(mu0)){
  ltt_temp = Ltt2[Ltt2$mu == mu0[i],]
  ltt_temp = rbind(ltt_temp, data.frame(t=15,Ex=ltt_temp$Ex[length(ltt_temp$Ex)]+1,mu=ltt_temp$mu[1]))
  pp = list(wt=c(ltt_temp$t[1],diff(ltt_temp$t)),E=rep(1,(length(ltt_temp$t)-1)),n=ltt_temp$Ex)
  phy = p2phylo(pp)
  stat[i] = ltt_stat(s$newick.extant,phy)
}

qplot(mu0,stat)

```

that look nice, let´s do it a bit more professional

```{r, lttvsmu}
time=proc.time()
stat=NULL
mu0 = seq(0.03,0.35,by=0.01)
phylo = s$newick.extant
for(i in 1:length(mu0)){
  mu = mu0[i]
  stat[i] = ltt_mu(mu,phylo)
}
print(proc.time()-time)
qplot(mu0,stat)
```


it looks fine, but 10 trees has too much variation, let´s repeat it for 100 trees


```{r, ex45, message=FALSE, eval=FALSE}
#ct = 15
#dt = 0.1
#grid = wt
Ltt = data.frame(t=cumsum(wt), Ex = (s$newick.extant.p)$n, mu=999)

for(i in 1:length(mu0)){
  mu = mu0[i]
  pars = c(l2[i],m2[i],k2[i]) 
  ltt = data.frame(t = cumsum(wt), Ex = expectedLTT(pars,drop.extinct = TRUE, grid=cumsum(wt))$Ex, mu=mu)
  Ltt = rbind(Ltt,ltt)
}
Ltt$mu = as.factor(Ltt$mu)
ggplot(data=Ltt, aes(x=t, y=Ex, colour = mu)) + geom_line() +   geom_point()
```


```{r, ex46, cache=FALSE, eval=FALSE}
ltt1 = Ltt[Ltt$mu == 999,]
diff_ltt = NaN
sq_diff = NaN
diff_mat = matrix(0,ncol=length(mu0),nrow=length(wt))
for(i in 1:length(mu0)){
 mu = mu0[i]
 ltt = Ltt[Ltt$mu == mu,]
 ltt$Ex = abs(ltt1$Ex-ltt$Ex)
 diff_mat[,i] = ltt$Ex
 diff_ltt[i] = sum(ltt$Ex)
 sq_diff[i] = sum((ltt1$Ex-ltt$Ex)^2)
 
}
diff_ltt_M = data.frame(mu = mu0, diff_ltt = diff_ltt)
choosed_mu = diff_ltt_M[diff_ltt_M$diff_ltt == min(diff_ltt_M$diff_ltt) ,]
choosed_mu$mu
b = matrix(0,ncol=length(mu0),nrow=length(wt))
for(i in 1:length(mu0)){
   b[i,] = diff_mat[i,]==rowMins(diff_mat)[i]
}
colSums(b)
MP[MP[,2] == choosed_mu$mu]
MP2[max(colSums(b)) == colSums(b)]
#qqplot()
```


(not working yet)

# Meta Analysis


Now we are prepared to estimate parameters ofr a set of (100) trees and see the distribution. 

The algorithm is: 

1. simulate tree and save MLE. Drop extinct species and save ltt
3. create a grid $\mu_g$ and run monte-carlo for every $\mu \in mu_g$, then get $(\lambda(\mu),mu,K(\mu)), \forall \mu \in \mu_g$ and the corresponding ltt
4. take the best $(\lambda(\mu),\mu,K(\mu))$ taking min ltt


```{r, supersim1, eval=FALSE}
ct = 15
dt = 0.1
#grid = seq(0,ct, by=dt)
n_it = 10
mu0 = seq(0.04,0.35,by=0.03)
n_trees = 10
MMP = matrix(nrow=n_it, ncol=3)
RMP = matrix(nrow=n_it, ncol=3)
for(j in 1:n_it){
  s = sim_phyl()
  p <- subplex(par = c(2,0.2,60), fn = llik, n = s$n, E = s$E, t = s$wt)$par
  MMP[j,] = p
  s2 = s$newick.extant.p
  wt = s2$wt
  grid = cumsum(wt)
  ltt1 = data.frame(t=grid, Ex = s2$n, mu=999)
  Ltt = ltt1
  MP = matrix(nrow=length(mu0), ncol=3)
  for(i in 1:length(mu0)){
    mu = mu0[i]
    trees = sim_srt(wt=wt, pars=c(p[1],mu,p[3]), parallel = F, n_trees = n_trees)
    pars = subplex(par = c(2,60), fn = llik_st , setoftrees = trees, mu = mu, impsam = FALSE)$par
    pars = c(pars[1],mu,pars[2])
    MP[i,] = pars
    ltt = data.frame(t = grid, Ex = expectedLTT(pars,drop.extinct = TRUE, grid=grid)$Ex, mu=mu)
    Ltt = rbind(Ltt,ltt)
  }
  diff_mat = matrix(0,ncol=length(mu0),nrow=length(wt))
  for(i in 1:length(mu0)){
   mu = mu0[i]
   ltt = Ltt[Ltt$mu == mu,]
   ltt$Ex = abs(ltt1$Ex-ltt$Ex)
   diff_mat[,i] = ltt$Ex
 }
 b = matrix(0,ncol=length(mu0),nrow=length(wt))
 for(i in 1:length(mu0)){
   b[i,] = diff_mat[i,]==rowMins(diff_mat)[i]
 }
 RMP[j,] = MP[max(colSums(b)) == colSums(b),]
}

```



```{r, eval=F}
RMP
MMP
```
