---
title: "Importance sampling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

```{r echo=FALSE, include=FALSE}
library(dmea)
library(diagram)
a=1
```
On this report we denote $D$ as the observed-part or extant-species, $+_i$ as the missing-part or extinct-species of the tree and $D^+$ is then the complete phylogenetic tree.  


The EM algorithms consists on two steps. First, we calculate the conditional expectation:

$$ Q(\theta|\theta^*) = E_{\theta^* } [log P(D^+|\theta) | D] $$

and then we performs the maximization:

$$ \theta ^{**} = argmax_{(\theta)} Q( \theta | \theta ^*) $$

Given that the calculation of the conditional expectation is really hard (if not impossible), we use an approximation, sampling complete-phylogenies under a montecarlo approach. This simulations should be sampled from (real density) 

$$ f_{\theta^*} (+_i | D) $$

But instead we sample it from 

$$ g_{\theta^*}(+_i | D) $$

To correct this we re-weigh the approximation of the expectation by importance scaling:

$$w_i =  \frac{f_{\theta^*} (+_i | D)}{g_{\theta^*}(+_i | D)} = \frac{f_i}{g_i}$$


Thus, the montecarlo approximation we are using has the form

$$ E_{\theta^* } [log P(D^+|\theta) | D] \approx \frac{1}{N} \sum^{N}_{i=1} log P(D_i^+ | \theta)  \frac{f_i}{g_i}$$


* $f_i$ is the density function (likelihood) of the complete phylogenetic tree. 

* to calculate $g_i$ we have 2 ways so far. The first one is related with the diagram above

```{r echo=FALSE}
par(mar=c(1,1,1,1))
openplotmat()
elpos<-coordinates (c(1,1,2,4))
fromto <- matrix(ncol=2,byrow=TRUE,data=c(1,2,2,3,2,4,4,7,4,8))
nr     <-nrow(fromto)
arrpos <- matrix(ncol=2,nrow=nr)
for (i in 1:nr) 
  arrpos[i,]<- straightarrow (to=elpos[fromto[i,2],],from=elpos[fromto[i,1],]
                              ,lwd=2,arr.pos=0.6,arr.length=0.5)
textellipse(elpos[1,],0.12,lab=paste("Update rates and",'\n',"draw next speciation time"),box.col="burlywood",shadow.col="darkgreen",shadow.size=0.005,cex=0.8)
textrect   (elpos[2,],0.15,0.05,lab=expression(paste('s', '<', Delta,t[i])),box.col="thistle",shadow.col="darkblue",shadow.size=0.005,cex=1.5)
textrect   (elpos[4,],0.15,0.05,lab=expression(paste('e', '<', t[p])),box.col="thistle",shadow.col="darkblue",shadow.size=0.005,cex=1.5)
textellipse(elpos[3,],0.1,0.1,lab=c("go to next bt"),box.col="aquamarine",shadow.col="red",shadow.size=0.005,cex=0.8)
#textellipse(elpos[3,],0.1,0.1,lab=c("other","term"),box.col="orange",shadow.col="red",shadow.size=0.005,cex=1.5)
textellipse(elpos[7,],0.1,0.1,lab=paste("add missing species","\n","and go to next bt"),box.col="aquamarine",shadow.col="red",shadow.size=0.005,cex=0.8)
textellipse(elpos[8,],0.1,0.1,lab=c("go to next bt"),box.col="aquamarine",shadow.col="red",shadow.size=0.005,cex=0.8)

dd <- c(0.0,0.025)
text(arrpos[2,1]+0.05,arrpos[2,2],"no")
text(arrpos[3,1]-0.05,arrpos[3,2],"yes")
text(arrpos[4,1]+0.05,arrpos[4,2]+0.05,"yes")
text(arrpos[5,1]-0.05,arrpos[5,2]+0.05,"no")
```

we multiply the corresponding probabilities to have every outcome, this is done on the piece of code bellow


```{r, eval=FALSE}
  if (t_spe < cwt){
      t_ext = rexp(1,mu0)
      t_ext = cbt + t_spe + t_ext
      if (t_ext < ct){
        up = update_tree(wt=wt,t_spe = (cbt + t_spe), t_ext = t_ext, E = E, n = n)
        E = up$E
        n = up$n
        wt = up$wt
        fake = FALSE
        prob[i] = dexp(t_ext,rate=mu0,log=TRUE) + dexp(t_spe,rate=s,log = TRUE)
      }else{
        prob[i] = pexp(q = ct,rate = mu0,lower.tail = F,log.p = TRUE) + dexp(t_spe,rate=s, log = TRUE)
        fake = TRUE
        i = i-1
      }
    }else{
      fake = FALSE
      prob[i] = pexp(q = cwt,rate = s,lower.tail = F,log.p = TRUE)
    }
```



Note that we are working with multiplication of many densities, which in the end are very small values. To avoid cutting most numbers to zero we consider the following fact:

$$ w_i = \frac{f_i}{g_i} = e^{log(f_i) - log(g_i)} = e^{loglik - \sum_j log(g_{i,j})} $$

and that is why on the code above the log of the densities is calculated. Considering that we calculate the weight in the following way



```{r, eval=FALSE}
f_n = -llik(b=pars,n=n,E=E,t=wt)
weight = exp(f_n-sum(prob))
return(list(wt=wt,E=E,n=n,weight=weight,L=L,g=prob,f_n=f_n))
```

another way to calculate $g$ is considering the probabilities of not having a new species on the interval $\Delta t_i$

$$ \int_0^{\Delta t_i} s_{\lambda_i}e^{-s_{\lambda_i} t}[1-e^{-\mu (r_i-t)}]  dt =  s_{\lambda_i}\int_0^{\Delta t_i}e^{-s_{\lambda_i} t}-e^{ t(\mu-s_{\lambda_i})-\mu r_i} = s_{\lambda_i}\int_0^{\Delta t_i}e^{-s_{\lambda_i} t}dt - s_{\lambda_i} e^{-\mu r_i} \int_0^{\Delta t_i}e^{  t(\mu-s_{\lambda_i})} dt $$
$$= 1-e^{-s_{\lambda_i} \Delta t} -\frac{s_{\lambda_i} e^{-\mu r_i}}{\mu - s_{\lambda_i}}[e^{\Delta t(\mu-s_{\lambda_i})}-1] $$
so we do it with this function

```{r, eval=FALSE}
convol <-function(wt,lambda,mu,remt){
  out = 1-exp(-lambda*wt)-lambda*exp(-mu*remt)/(mu-lambda)*(exp(wt*(mu-lambda))-1)
  return(out)
}
```

and we add this option to the code

```{r, eval=FALSE}
 if (t_spe < cwt){
      t_ext = rexp(1,mu0)
      t_ext = cbt + t_spe + t_ext
      if (t_ext < ct){
        up = update_tree(wt=wt,t_spe = (cbt + t_spe), t_ext = t_ext, E = E, n = n)
        E = up$E
        n = up$n
        wt = up$wt
        fake = FALSE
        prob[i] = dexp(t_ext,rate=mu0,log=TRUE) + dexp(t_spe,rate=s,log = TRUE)
      }else{
        prob[i] = pexp(q = ct,rate = mu0,lower.tail = F,log.p = TRUE) + dexp(t_spe,rate=s,log = TRUE)
        if(v2){ prob[i] = log(convol(wt = t_spe,lambda = s,mu = mu,remt = ct-cbt))}
        fake = TRUE
        i = i-1
      }
    }else{
      fake = FALSE
      prob[i] = pexp(q = cwt,rate = s,lower.tail = F,log.p = TRUE)
      if(v2){prob[i] = log(convol(wt = t_spe,lambda = s,mu = mu,remt = ct-cbt))}
    }
```

### Implementation

To have a better idea on how this implementation look like we do some simple observations. 

We start simulating a whole tree

```{r}
s <- sim_phyl(seed=2)
plot(s$newick)
```  

Them we drop extinct species

```{r}
s2 <- drop.fossil(s$newick)
plot(s2)
```

we transform phylo format into branching-times, number of species, and topology vectors

```{r}
s2 <- phylo2p(s2)
```

Now we simulate a complete tree (extinct+extans species) based on the observed tree (extant species only) 

```{r, warning=FALSE}
s3 <- rec_tree(wt=s2$t)
```

and we observe the calculated weight for this tree

```{r}
s3$weight
```

it seems very small

This is a random process, if we do it again we end with another complete-tree and its corresponding weight, we do it for 30 iterations just to have an idea of the variability of the weight


```{r, warning=FALSE} 
w=0
for(i in 1:100){
st3 <- rec_tree(wt=s2$t)
w[i] = st3$weight 
}
print(w)
```

```{r}
w = w - max(w)
print(w)
w = exp(w)
print(w)
```

So, a first observation, <span style="color:red">  the values of w are really small and varies a lot. That is amaking a lot of numerical inestabilities. </span>



```{r, warning=FALSE} 
boxplot(w)
summary(w)
```

<span style="color:red">  we can also see how few trees has almost the whole weight and defines the outcome  </span>



```{r, warning=FALSE} 
#print(w/max(w))
```
***

now the same but with the second way

```{r, warning=FALSE}
s3 <- rec_tree(wt=s2$t,v2=T)
```

and we observe the calculated weight for this tree

```{r}
s3$weight
```

it seems very small

This is a random process, if we do it again we end with another complete-tree and its corresponding weight, we do it for 30 iterations just to have an idea of the variability of the weight


```{r, warning=FALSE} 
w=0
for(i in 1:100){
st3 <- rec_tree(wt=s2$t,v2=T)
w[i] = st3$weight 
}
print(w)
```

```{r}
w = w - max(w)
print(w)
w = exp(w)
print(w)
```

So, a first observation, <span style="color:red">  the values of w are really small and varies a lot. That is amaking a lot of numerical inestabilities. </span>

```{r, warning=FALSE} 
boxplot(w)
summary(w)
```

<span style="color:red">  we can also see how few trees has almost the whole weight and defines the outcome  </span>

we try dividing by the max, but it does not changes 

```{r, warning=FALSE}
#boxplot(w/max(w))
#summary(w/max(w))
```


```{r, warning=FALSE} 
#print(w/max(w))
```

