---
title: "Importance sampling"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, include=FALSE}
library(dmea)
```
On this report we denote $D$ as the observed-part or extant-species, $+_i$ as the missing-part or extinct-species of the tree and $D^+$ is then the complete phylogenetic tree.  


The EM algorithms consists on two steps. First, we calculate the conditional expectation:

$$ Q(\theta|\theta^*) = E_{\theta^* } [log P(D^+|\theta) | D] $$

and then we performs the maximization:

$$ \theta ^{**} = argmax_{(\theta)} Q( \theta | \theta ^*) $$

Given that the calculation of the conditional expectation is really hard (if not impossible), we use an approximation, sampling complete-phylogenies under a montecarlo approach. This simulations should be sampled from (real density) 

$$ f_{\theta^*} (+_i | D) $$

But instead we sample it from 

$$ g_{\theta^*}(+_i | D) $$

To correct this we re-weigh the approximation of the expectation by importance scaling:

$$w_i =  \frac{f_{\theta^*} (+_i | D)}{g_{\theta^*}(+_i | D)} $$


Thus, the montecarlo approximation we are using has the form

$$ E_{\theta^* } [log P(D^+|\theta) | D] \approx \frac{1}{N} \sum^{N}_{i=1} log P(D_i^+ | \theta)  \frac{f_{\theta^*} (+_i | D)}{g_{\theta^*}(+_i | D)}$$

Moreover, 

$$ f_{ \theta^* } (+_i | D) = \frac{f_{\theta^*} (+_i \cap D)}{f_{\theta^*} (D)} = c \cdot f_{\theta^*} (D^+) $$ 

where $c = \frac{1}{f_{\theta^*} (D)}$ does not depends on the $i$-iteration of the EM algorithm. And $f_{\theta^*} (D^+)$ is the density function of the complete phylogenetic tree. 

Thus, we calculate $g_{\theta^*}(+_i | D)$ within the ```rec_tree``` function which calculates the density $g$ at the same time that simulates the missing species. The part of the code which does that is the one below

```{r, eval=FALSE}
t_spe = rexp(1,s)
prob = prob*dexp(t_spe,rate=s)  #adding the speciation event to g
if (t_spe < cwt){
  t_ext = rexp(1,mu0)
  prob = prob*dexp(t_ext,rate=mu0) #adding extinction event to g
```

and the same code calculates $f_{\theta^*} (D^+)$ just taking the exponential of the -log-likelihood function the the complete reconstructed tree. After that we calculate the weight:

```{r, eval=FALSE}
f_n = exp(-llik(b=pars,n=n,E=E,t=wt))
weight = f_n/prob2
```

##### Implementation

To have a better idea on how this implementation look like we do some simple observations. 

We start simulating a whole tree

```{r}
st <- sim_phyl()
plot(st$newick)
```  

Them we drop extinct species

```{r}
st2 <- drop.fossil(st$newick)
plot(st2)
```

we transform phylo format into branching-times, number of species, and topology vectors

```{r}
st2 <- phylo2p(st2)
```

Now we simulate a complete tree (extinct+extans species) based on the observed tree (extant species only) 

```{r, warning=FALSE}
st3 <- rec_tree(wt=st2$t)
```

and we observe the calculated weight for this tree

```{r}
st3$weight
```

it seems very small, we can see also the values of $f$ and $g$ to have an idea of the order of them 

```{r}
st3$f_n
st3$prob
```


This is a random process, if we do it again we end with another complete-tree and its corresponding weight, we do it for 30 iterations just to have an idea of the variability of the weight


```{r, warning=FALSE} 
for(i in 1:30){
st3 <- rec_tree(wt=st2$t)
print(st3$weight)
}
```
So, a first observation, <span style="color:red">  THE WEIGHT VALUES ARE REALLY SMALL AND VARIES A LOT WITH DIFFERENT TREES. DO YOU THINK IS THAT OK? </span>


***


#### Estimations

Moreover, we run estimations with importance sampling now (we would love that the estimations of $\mu$ would not be underestimated)

```{r sim1}

p = proc.time()
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)

n_sim = 1000
n_trees = 10
MP = matrix(nrow=n_sim,ncol=3)
RP = matrix(nrow=n_sim,ncol=3)

ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% {  
  set.seed(i)
  sim_est(n_trees=n_trees,impsam=TRUE)
}
for (i in 1:n_sim){
  RP[i,] = ests[,(2*i-1)]
  MP[i,] = ests[,2*i]
}
stopCluster(cl)
print(proc.time()-p)
```


first we look at lambda:

```{r plot1}
par_est_vis(P=MP,par=1,PR=RP)
```

We can see now that there are few trees that could not get a proper estimation, [This does not happens on the estimations without importance sampling](http://rpubs.com/franciscorichter/231599).

To visualize better the other estimations we just remove that ones

```{r}
par_est_vis(P=MP[MP[,1]<2,],par=1,PR=RP[MP[,1]<2,])
```

We have some outliers that we did not had before. Let´s see how are the estimations of $\mu$

```{r}
par_est_vis(P=MP[MP[,1]<3,],par=3,PR=RP[MP[,1]<3,])
```

It seems they are still biased as [before](http://rpubs.com/franciscorichter/231599)



#### More code
In case is useful, the function used to simulate the extinct part of the recontructe tree is the one below

```{r, eval=F}
rec_tree <- function(wt, pars=c(0.8,0.0175,0.1), model='dd'){
  lambda0 = pars[1]
  mu0 = pars[3]
  K = (lambda0-mu0)/pars[2]
  n = 1:length(wt)
  i = 1
  tails = tail(n,n=1)
  E = rep(1,(length(wt)-1))
  fake = FALSE
  ct = sum(wt)
  prob = 1
  while(i < length(wt)){
    N = n[i]
    if(model == "dd"){  # diversity-dependence model
      lambda = max(0,lambda0 - (lambda0-mu0)*N/K)
      mu = mu0
      lambda = rep(lambda,N)
      mu = rep(mu,N)
    }
    if(model == 'cr'){ # constant-rate model
      lambda = rep(lambda0,N)
      mu = rep(mu0,N)
    }
    s = sum(lambda)
    if(s==0){
      #print('s=0')
      break
    }
    if(fake){ # in case there was an speciation but not extinction previously
      cwt = cwt - t_spe
      cbt = cbt + t_spe
    }
    else{
      cwt = wt[i]
      cbt = cumsum(wt)[i] - cwt
    }
    t_spe = rexp(1,s)
    prob = prob*dexp(t_spe,rate=s)
    if (t_spe < cwt){
      t_ext = rexp(1,mu0) # this is not as general as trees with trait-dependance species yet,
      prob = prob*dexp(t_ext,rate=mu0)
      t_ext = cbt + t_spe + t_ext
      if (t_ext < ct){
        up = update_tree(wt=wt,t_spe = (cbt + t_spe), t_ext = t_ext, E = E, n = n)
        E = up$E
        n = up$n
        wt = up$wt
        fake = FALSE
      }else{
        fake = TRUE
        i = i-1
      }
    }else{
      fake = FALSE
    }
    i = i+1
  }
  L = create_L(wt,E)
  f_n = exp(-llik(b=pars,n=n,E=E,t=wt))
  weight = f_n/prob
  return(list(wt=wt,E=E,n=n,weight=weight,L=L,prob=prob,f_n=f_n))
}
```

<!--

```{r plot2}
par_est_vis(P=MP[MP[,1]<3,],par=1,PR=RP[MP[,1]<3,])
```
 --> 




<!--
Where, in the case of  $f_{\theta^*} (+_i | D)  = \prod_i P_{ms_i}$, and $P_{ms_i}$ corresponds to the probability for the missing species number $i$. That is 

$$ P_{ms_i} = \displaystyle\underbrace{\left(\sum \lambda\right)_{i} e^{-t_i(\sum \lambda)_i} \frac{1}{n_i}}_{speciation} \displaystyle\underbrace{\frac{\mu e^{-t_i \mu}}{1-e^{-\mu rT}}}_{extinction} $$

Then, for the case of DD we have 

$$ P_{ms_i} = n_i(\lambda_0-\beta n_i)  e^{-t_i(n_i(\lambda_0-\beta n_i))} \frac{1}{n_i} \frac{\mu e^{-t_i^{ext} \mu}}{1-e^{-\mu(ct-bt_i)}} $$
 --> 
