---
title: "December 14, 2016. updated 27 jan"
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(dmea)
```


Following previous notation, if we call $l(\theta;D^+)$ the -log-likelihood function of the complete phylogenetic tree we have a function that looks like this

$$ l(\theta;D^+) = \sum n_i t_i - log(\rho_i) $$

 If we focuss only on the influence of extinctions, and we assume constant extinction rate $\mu$, we can write this equation on the following way

$$ l(\theta;D^+) = \sum (n_i t_i-E_i log(\lambda_i)) \color{blue}{ - (\#extinctions \cdot log(\mu)) } $$


Then, if $\mu < 1$ we would have that $l(\theta;D^+)$ increases with the number of extinctions the tree has (note that we want to minimaze $l(\theta;D^+)$). 

This show us two anoying facts:

  * Trees with less missing species have larger likelihood (if $\mu$<1)
  
  * The model is time scale-dependent (the inflextion point $\mu$<1 as different meanings with different time-scales).
  


***
The following are just simples exercises that helps on understanding the relationship beween the number of extinctions and the likelihood of a tree, in the context of importance sampling. 


We would like to see the relationship between the importance weight and the extinctions on a tree. To do that, given an observed tree we reconstruct 100 (extinct included) complete trees.  

```{r, warning=FALSE}

num_miss = NULL
imp_scal = NULL
#g = NULL
#f = NULL
seed=2
st = sim_phyl(seed=seed)
extant = phylo2p(drop.fossil(st$newick))
pars=c(0.8,0.0175,0.1)
S = sim_srt(wt=extant$t, pars=pars, parallel=T, n_trees=1000)
for(i in 1:1000){
  imp_scal[i] = S[[i]]$weight
  #num_miss[i] = (length(rec$wt)-length(extant$t))/2
  #g[i] = rec$prob
  #f[i] = rec$f_n
}
#plot(num_miss,log(imp_scal))
boxplot(imp_scal)
```

```{r}
imp_scal
summary(imp_scal)
#lm(log(imp_scal) ~ num_miss)

```

