---
title: "December 14, 2016"
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(dmea)
```


Following previous notation, if we call $l(\theta;D^+)$ the -log-likelihood function of the complete phylogenetic tree we have a function that looks like this

$$ l(\theta;D^+) = \sum n_i t_i - log(\rho_i) $$

 If we focuss only on the influence of extinctions, and we assume constant extinction rate $\mu$, we can write this equation on the following way

$$ l(\theta;D^+) = \sum (n_i t_i-E_i log(\lambda_i)) \color{blue}{ - (\#extinctions \cdot log(\mu)) } $$


Then, if $\mu < 1$ we would have that $l(\theta;D^+)$ increases with the number of extinctions the tree has (note that we want to minimaze $l(\theta;D^+)$). 

This show us two anoying facts:

  * Trees with less missing species have larger likelihood (if $\mu$<1)
  
  * The model is time scale-dependent (the inflextion point $\mu$<1 as different meanings with different time-scales).
  


***
The following are just simples exercises that helps on understanding the relationship beween the number of extinctions and the likelihood of a tree, in the context of importance sampling. 


We would like to see the relationship between the importance weight and the extinctions on a tree. To do that, given an observed tree we reconstruct 100 (extinct included) complete trees.  

```{r, warning=FALSE}
num_miss = NULL
imp_scal = NULL
g = NULL
f = NULL
st = sim_phyl()
extant = phylo2p(drop.fossil(st$newick))
for(i in 1:100){
  rec = rec_tree(wt=extant$t)
  imp_scal[i] = rec$weight
  num_miss[i] = (length(rec$wt)-length(extant$t))/2
  g[i] = rec$prob
  f[i] = rec$f_n
}
plot(num_miss,log(imp_scal))
```

```{r lm1}
imp_scal
#lm(log(imp_scal) ~ num_miss)

```

Moreover, the number of real missing species is 
```{r}
number_missing(st)
```


and the set number of the simulated species are
```{r}
summary(num_miss)
```

the relationship between between f ang g as described [here](http://rpubs.com/franciscorichter/231493)

```{r}
plot(log(f),log(g))
```


the relationship between f and number of species
```{r}
plot(num_miss,log(f))
```


the relationship between g and number of species
```{r}
plot(num_miss,log(g))
```



***

The plots above are interesting, let's go one step back and see the relationship between the likeihood of a tree and the number of species (extinct and extant)


We simulate trees, count number of extinct species and the corresponding -log-likelihood function

```{r}

n_extinct = NULL
ll = NULL
sum_spec = NULL
sum_ext = NULL
for(i in 1:100){
  s = sim_phyl(seed=TRUE,seed_val = i)
  ll[i] = llik(b=c(0.8,0.0175,0.1),n=s$n,E=s$E,t=s$t)
  n_extinct[i] = number_missing(s)
  sum_spec[i] = sum(s$E)
  sum_ext[i] = sum(1-s$E)
}
plot(n_extinct,ll)
```



```{r}
lm(ll~n_extinct)
```

