---
title: "Report"
author: "F. Richter"
date: "December 7th, 2015"
output: rmarkdown::tufte_handout
---

# Introduction 

In this report an MLE estimation of the constant rates model is calculated. The general model, with non constant birth-death rates, is also discussed (I could not find an analytical MLE solution for that case). 

In addition, simulated data and an intent of finding a numerical solution were performed. 


# Analitycal analysis

  We define, 
    

\begin{itemize}

\item  $\beta$: Speciation rate.
  
\item  $\delta$ : Extinction rate.
  
\item  $t_i$: Time when the event $i$ (speciation or extinction) occurs.

\item  $x_i$: Number of species on time $t_i$. 
  
\end{itemize}


The probability of a speciation ocurrs on time $t$ is given by the exponential probability distribution 
\begin{figure}
    $$ P_\beta(T=t;\beta) = -\beta e^{-\beta t} $$
   % \caption{Probability of a speciation occurs on time $t$}
\end{figure}     
whereas the probabilty of an extinction occurs on time $t$ is
\begin{figure}
    $$ P_\delta(T=t;\delta) = -\delta e^{-\delta t} $$
      %  \caption{Probability of a speciation ocurrs on time $t$}
\end{figure}   
Moreover, given the waiting time $t_1$ the probability of an speciation event occurs is $\beta / (\beta + \delta)$ and the probability an expeciation happens on $t_i$ is $\delta / (\beta + \delta)$ 

\begin{marginfigure}
    $$ P(Birth;\emph{waiting time}) = \frac{\beta}{\beta + \delta} $$
    $$ P(Death;\emph{waiting time}) = \frac{\delta}{\beta + \delta} $$
    \caption{Probability of speciation and extintions event.}
\end{marginfigure}  

Thus, a possible likelihood function corrisponding to the phylogenic tree would be 

\begin{figure}

$$L = \displaystyle\prod_{i=1}^n (\beta + \delta) x_i e^{-(\beta + \delta) x_i t_i}  \left( \frac{\beta}{\beta+\delta} \right)^{b_i} \left( \frac{\delta}{\beta+\delta} \right)^{d_i}$$

\end{figure}


where $b_i$ is the number of speciations on time $t_i$ and $d_i$ is the number of extinctions on time $t_i$. 

Thus, the loglikelihood function is 

\begin{figure}

$$lL = \displaystyle\sum_{i=1}^n ln(\beta + \delta) + ln(x_i)  -(\beta + \delta) x_i t_i + b_i(ln(\beta)-ln(\beta+\delta)) + d_i(ln(\delta)-ln(\beta+\delta)) $$

\end{figure}


Finally, to maximize the likelihood we solve the equations 

\begin{equation}
\frac{\partial lL}{\partial \beta} = 0 ; \qquad
\frac{\partial lL}{\partial \delta} = 0
\label{mle}
\end{equation}


The analytical solution corresponding to equations (\ref{mle}) are 

\begin{figure}

$$\beta = \frac{N\sum b_i}{\sum x_it_i (\sum b_i + \sum d_i)} $$
$$\delta = \frac{N\sum d_i}{\sum x_it_i (\sum b_i + \sum d_i)}  $$

\end{figure}



This is an useful result as a first step, however, since we assume $\beta$ and $\delta$ are constant numbers, this is not yet a realistic case. We would be interested in the case when the speciation and extinction parameters depends on covariates, for instance 




$$log(\beta) = \theta_0 + \theta_1 \emph{body size} + \theta_2 \emph{ave. temp.} $$






\begin{marginfigure}

$$log(\beta) = A\theta $$
$$log(\delta) = A\phi  $$

\caption{An option is to model the parameters as a GLM}
\end{marginfigure}


We consider the simple case 

$$\beta = e^{\theta_0 + \theta_1 a}$$
$$\delta = e^{\phi_0 + \phi_1 a}  $$

and we call it the model M2.

\begin{marginfigure}

$$\frac{\partial\beta}{\partial \theta_0} = e^{\theta_0 + \theta_1 a}  $$
$$\frac{\partial\beta}{\partial \theta_1} = ae^{\theta_0 + \theta_1 a}  $$

\end{marginfigure}

\begin{marginfigure}

$$\frac{\partial\delta}{\partial \theta_0} = e^{\phi_0 + \phi_1 a}  $$
$$\frac{\partial\delta}{\partial \theta_1} = ae^{\phi_0 + \phi_1 a}  $$

\end{marginfigure}

For MLE calculation, by the chain rule we have 


$$ \frac{\partial lL}{\partial \theta_0} = \frac{\partial lL}{\partial \beta} \frac{\partial \beta}{\partial \theta_0} = e^{\theta_0 + \theta_1 a} \frac{\partial lL}{\partial \beta} ; $$
$$ \frac{\partial lL}{\partial \theta_1} = \frac{\partial lL}{\partial \beta} \frac{\partial \beta}{\partial \theta_0} = ae^{\theta_0 + \theta_1 a} \frac{\partial lL}{\partial \beta}, $$

and similarly 

$$ \frac{\partial lL}{\partial \phi_0}  = e^{\phi_0 + \phi_1 a} \frac{\partial lL}{\partial \delta} ; $$
$$ \frac{\partial lL}{\partial \phi_1}  = ae^{\phi_0 + \phi_1 a} \frac{\partial lL}{\partial \delta}.$$

Thus, if $a\neq 0$, the system 

\begin{equation}
\frac{\partial lL}{\partial \theta_0} = 0 ; \quad
\frac{\partial lL}{\partial \theta_1} = 0 ; \quad
\frac{\partial lL}{\partial \phi_0} = 0 ; \quad
\frac{\partial lL}{\partial \phi_1} = 0 
\label{mle2}
\end{equation}

is equivalent to (\ref{mle}) and we have two equations but four parameters. 

# Simulations

We simulate data from the M2 model:

```{r, warning=FALSE}
nT=15
theta=c(1,2)
phi=c(0,1)
set.seed(123)
dat<-NULL
AA <- NULL
AA <- list()
A=rexp(1,4)
dat[[1]]<-list(tm=0,x=1,A=A)
X=0
X[1]=1
tm=0
D=0
B=0
Tm=tm
AA[[1]]=A
i<-2
stm <- 0
while (i<=nT & dat[[i-1]]$x>0){
  beta<-theta[1]+theta[2]*dat[[i-1]]$A
  delta<-phi[1]+phi[2]*dat[[i-1]]$A
  prev.x=dat[[i-1]]$x  
  tm<-rexp(1,(sum(beta)+sum(delta)))
  stm <- stm+tm
  prob<-c(delta,beta)/(sum(beta)+sum(delta))
  BD<-sample(2*prev.x,1,prob=prob)
  if(BD<=prev.x){
    x = prev.x-1
    A = dat[[i-1]]$A[-BD]
    D[i]=1
    B[i]=0
  } else {
    x = prev.x + 1
    A = c(dat[[i-1]]$A,dat[[i-1]]$A[BD-prev.x]*rgamma(1,100,100))
    D[i]=0
    B[i]=1
  }
  dat[[i]]<-list(tm=stm,x=x,A=A)
  X[i]=x
  Tm[i]=tm
  AA[[i]]=A
  
  i<-i+1
}

```

To see the iterations we type
```{r}
dat
```
\newpage
That is

```{r, results='asis', xtable.margin = TRUE}
it=length(dat)
tree <- data.frame(time=it,x=2)
for (i in 1:it){
  tree[i,] = c(t=dat[[i]]$tm,x=dat[[i]]$x)
}
library(xtable)
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(tree, caption = "Table with simulated waiting times and number of species")
```


We try an MLE calculation maximizing the Likelihood function, but for different initial values the estimated parameters are very different and most of them does not make any sense. 


```{r, warning=FALSE}
fn <- function(theta) {
  beta=0
  delta=0
  for(i in 1:nT){
    beta[i]=theta[1]+sum(theta[2]*AA[[i]])
    delta[i]=theta[3]+sum(theta[4]*AA[[i]])
  }
  sum(log(beta+delta)+log(x)-(beta+delta)*x*Tm+D*(log(beta)-log(beta+delta))+B*(log(delta)-log(beta+delta)))
}

nlm(fn, theta <- c(0.5,0.1,0.7,0.8), hessian=TRUE)
```




