---
title: "Last MCEM step"
author: "F.R.M."
date: "November 28, 2016"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, include=FALSE}
library(dmea)

```

## A first exercise

1. The normal one...

```{r eval=FALSE}

p = proc.time()
n_sim = 1000
n_trees = 100
MP = matrix(nrow=n_sim, ncol=3)
RP = matrix(nrow=n_sim, ncol=3)
for(j in 1:n_sim){
  st = sim_phyl(seed=j)
  p <- subplex(par = c(8,0.175,0.9),fn = llik,n = st$n, E = st$E, t = st$t)$par
  RP[j,] = p
  sit = dmea::drop.fossil(st$newick)
  sit = phylo2p(sit)
  trees = sim_srt(tree=sit, pars=p, parallel = F, n_trees = 100)
  pars = subplex(par = c(8,0.175,0.9), fn = llik_st , setoftrees = trees, impsam = FALSE)$par
  MP[j,] = pars
}
par_est_vis(P=MP,par=1,PR=RP)
par_est_vis(P=MP,par=2,PR=RP)
par_est_vis(P=MP,par=3,PR=RP)
print(proc.time()-p)
```
This one took toooo much time..

2. Now the same, but trying to run it in parallel

```{r sim2}
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)


n_sim = 100
n_trees = 10
MP = matrix(nrow=n_sim,ncol=3)
RP = matrix(nrow=n_sim,ncol=3)
p = proc.time()
ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% sim_est(n_trees=n_trees)
print(proc.time()-p)
for (i in 1:n_sim){
  RP[i,] = ests[,(2*i-1)]
  MP[i,] = ests[,2*i]
}
par_est_vis(P=MP,par=1,PR=RP)
par_est_vis(P=MP,par=2,PR=RP)
par_est_vis(P=MP,par=3,PR=RP)



stopCluster(cl)
```


3. Now the same, but more simulations in parallel


```{r sim3}
p = proc.time()
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)

n_sim = 1000
n_trees = 10
MP = matrix(nrow=n_sim,ncol=3)
RP = matrix(nrow=n_sim,ncol=3)

ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% {  
  set.seed(i)
  sim_est(n_trees=n_trees)
}
for (i in 1:n_sim){
  RP[i,] = ests[,(2*i-1)]
  MP[i,] = ests[,2*i]
}
par_est_vis(P=MP,par=1,PR=RP)
par_est_vis(P=MP,par=2,PR=RP)
par_est_vis(P=MP,par=3,PR=RP)
stopCluster(cl)
print(proc.time()-p)
```

```{r}
summary(MP)
summary(RP)
```

4. Now more trees for each iteration 

```{r sim4}
p = proc.time()
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)

n_sim = 1000
n_trees = 100
MP2 = matrix(nrow=n_sim,ncol=3)
RP2 = matrix(nrow=n_sim,ncol=3)

ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% {  
  set.seed(i)
  sim_est(n_trees=n_trees)
}
for (i in 1:n_sim){
  RP2[i,] = ests[,(2*i-1)]
  MP2[i,] = ests[,2*i]
}
par_est_vis(P=MP2,par=1,PR=RP2)
par_est_vis(P=MP2,par=2,PR=RP2)
par_est_vis(P=MP2,par=3,PR=RP2)
stopCluster(cl)
print(proc.time()-p)
```

```{r}
summary(MP2)
summary(RP2)
```

This is strange, with 100 trees for monte-carlo simulation we introduce some biases to lambda and beta that with 10 trees MC we did not have. 

5. Does it makes sense to run 1 tree monte carlo?

```{r sim5}
p = proc.time()
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)

n_sim = 1000
n_trees = 1
MP2 = matrix(nrow=n_sim,ncol=3)
RP2 = matrix(nrow=n_sim,ncol=3)

ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% {  
  set.seed(i)
  sim_est(n_trees=n_trees)
}
for (i in 1:n_sim){
  RP2[i,] = ests[,(2*i-1)]
  MP2[i,] = ests[,2*i]
}
par_est_vis(P=MP2,par=1,PR=RP2)
par_est_vis(P=MP2,par=2,PR=RP2)
par_est_vis(P=MP2,par=3,PR=RP2)
stopCluster(cl)
print(proc.time()-p)
```







6. Now importance sampling


```{r sim6}
p = proc.time()
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)

n_sim = 1000
n_trees = 10
MP = matrix(nrow=n_sim,ncol=3)
RP = matrix(nrow=n_sim,ncol=3)

ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% {  
  set.seed(i)
  sim_est(n_trees=n_trees, impsam = TRUE)
}
for (i in 1:n_sim){
  RP[i,] = ests[,(2*i-1)]
  MP[i,] = ests[,2*i]
}
par_est_vis(P=MP,par=1,PR=RP)
par_est_vis(P=MP,par=2,PR=RP)
par_est_vis(P=MP,par=3,PR=RP)
stopCluster(cl)
print(proc.time()-p)
```

```{r}
summary(MP)
summary(RP)
```

7. Now more trees for each iteration 

```{r sim7}
p = proc.time()
no_cores <- detectCores()- 1 
cl <- makeCluster(no_cores)
registerDoParallel(cl)

n_sim = 1000
n_trees = 10
MP2 = matrix(nrow=n_sim,ncol=3)
RP2 = matrix(nrow=n_sim,ncol=3)

ests <- foreach(i = 1:n_sim, .combine=data.frame,.packages='dmea') %dopar% {  
  set.seed(i)
  sim_est(n_trees=n_trees,impsam=TRUE)
}
for (i in 1:n_sim){
  RP2[i,] = ests[,(2*i-1)]
  MP2[i,] = ests[,2*i]
}
par_est_vis(P=MP2,par=1,PR=RP2)
par_est_vis(P=MP2,par=2,PR=RP2)
par_est_vis(P=MP2,par=3,PR=RP2)
stopCluster(cl)
print(proc.time()-p)
```
