---
title: "Reconstructions methods comparison"
output: html_document
---


````{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, include=FALSE,cache=FALSE}
library(dmea)
library(reshape2)
library(nLTT)
```

On this report we aim to compare the two main reconstruction methods and set up a standar tool for comparison for other purposes. 

To do that we consider the following statistics:

* nLTT statistic: Seems to be the most important one in our context
* Gamma statistic: A value <0 indicates internal nodes close to expected tips on pure-birth model and the other way around
* Phylogenetic Diversity statistic: Measures the phylogenetic diversity

## Method 1 (standard)

```{r sim1}
pp=proc.time()
s = sim_phyl(seed=2)
s2 = phylo2p(drop.fossil(s$newick))
n_trees=5000
phylos=vector('list',length=n_trees)
nLTT = vector(length=n_trees)
gam = vector(length=n_trees)
phylos2=vector('list',length=n_trees)
nLTT2 = vector(length=n_trees)
gam2 = vector(length=n_trees)
dt=0.01
for(i in 1:n_trees){
  r = rec_tree(wt=s2$t)
  r1 = p2phylo(r)
  phylos[[i]] = read.tree(text=r1)
  nLTT[i] = nLTTstat(s$newick, phylos[[i]], distance_method = "abs")
  gam[i] = gammaStat(phylos[[i]])
  #
  r = rec_tree(wt=s2$t,rec_method=2)
  r2 = p2phylo(r)
  phylos2[[i]] = read.tree(text=r2)
  nLTT2[i] = nLTTstat(s$newick, phylos2[[i]], distance_method = "abs")
  gam2[i] = gammaStat(phylos2[[i]])
}
nltt_values <- get_nltt_values(phylos, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)

dat = data.frame(nltt=c(nLTT,nLTT2),gamma = c(gam,gam2),method=c(rep('method 1',length(nLTT)),rep('method 2',length(nLTT))))
ggplot(dat, aes(x=nltt, fill=method)) + geom_histogram()
ggplot(dat, aes(x=gamma, fill=method)) + geom_histogram()
print(pp-proc.time())
```

```{r}
nltt_values <- get_nltt_values(phylos2, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)
```



```{r sim2}
s = sim_phyl(seed=2)
s2 = phylo2p(drop.fossil(s$newick))
n_trees=5000
phylos=vector('list',length=n_trees)
nLTT = vector(length=n_trees)
gam = vector(length=n_trees)
phylos2=vector('list',length=n_trees)
nLTT2 = vector(length=n_trees)
gam2 = vector(length=n_trees)
dt=0.001
for(i in 1:n_trees){
  r = rec_tree(wt=s2$t)
  r1 = p2phylo(r)
  phylos[[i]] = read.tree(text=r1)
  nLTT[i] = nLTTstat(s$newick, phylos[[i]], distance_method = "abs")
  gam[i] = gammaStat(phylos[[i]])
  #
  r = rec_tree(wt=s2$t,rec_method=2)
  r2 = p2phylo(r)
  phylos2[[i]] = read.tree(text=r2)
  nLTT2[i] = nLTTstat(s$newick, phylos2[[i]], distance_method = "abs")
  gam2[i] = gammaStat(phylos2[[i]])
}
nltt_values <- get_nltt_values(phylos, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)

dat = data.frame(x=c(nLTT,nLTT2),method=c(rep('method 1',length(nLTT)),rep('method 2',length(nLTT))))
ggplot(dat, aes(x=x, fill=method)) + geom_histogram(alpha=0.2, position="identity")

```

```{r}
nltt_values <- get_nltt_values(phylos2, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)
```

```{r}
dat = data.frame(nltt=c(nLTT,nLTT2),gamma = c(gam,gam2),method=c(rep('method 1',length(nLTT)),rep('method 2',length(nLTT))))
ggplot(dat, aes(x=nltt, fill=method)) + geom_histogram(alpha=0.5)
ggplot(dat, aes(x=gamma, fill=method)) + geom_histogram(alpha=0.3)
ggplot(dat, aes(x=gamma, fill=method)) + geom_histogram(alpha=0.5, position='identity')
```

# more sim
```{r sim3}
s = sim_phyl(seed=2)
s2 = phylo2p(drop.fossil(s$newick))
n_trees=10000
phylos=vector('list',length=n_trees)
nLTT = vector(length=n_trees)
gam = vector(length=n_trees)
phylos2=vector('list',length=n_trees)
nLTT2 = vector(length=n_trees)
gam2 = vector(length=n_trees)
dt=0.001
for(i in 1:n_trees){
  r = rec_tree(wt=s2$t)
  r1 = p2phylo(r)
  phylos[[i]] = read.tree(text=r1)
  nLTT[i] = nLTTstat(s$newick, phylos[[i]], distance_method = "abs")
  gam[i] = gammaStat(phylos[[i]])
  #
  r = rec_tree(wt=s2$t,rec_method=2)
  r2 = p2phylo(r)
  phylos2[[i]] = read.tree(text=r2)
  nLTT2[i] = nLTTstat(s$newick, phylos2[[i]], distance_method = "abs")
  gam2[i] = gammaStat(phylos2[[i]])
}
nltt_values <- get_nltt_values(phylos, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)

dat = data.frame(x=c(nLTT,nLTT2),method=c(rep('method 1',length(nLTT)),rep('method 2',length(nLTT))))
ggplot(dat, aes(x=x, fill=method)) + geom_histogram(alpha=0.2, position="identity")

```


```{r}
nltt_values <- get_nltt_values(phylos2, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)
```

# more sim
```{r sim4}
pp = proc.time()
s = sim_phyl(seed=2)
s2 = phylo2p(drop.fossil(s$newick))
n_trees=1000
phylos=vector('list',length=n_trees)
nLTT = vector(length=n_trees)
gam = vector(length=n_trees)
phylos2=vector('list',length=n_trees)
nLTT2 = vector(length=n_trees)
gam2 = vector(length=n_trees)
dt=0.0001
for(i in 1:n_trees){
  r = rec_tree(wt=s2$t)
  r1 = p2phylo(r)
  phylos[[i]] = read.tree(text=r1)
  nLTT[i] = nLTTstat(s$newick, phylos[[i]], distance_method = "abs")
  gam[i] = gammaStat(phylos[[i]])
  #
  r = rec_tree(wt=s2$t,rec_method=2)
  r2 = p2phylo(r)
  phylos2[[i]] = read.tree(text=r2)
  nLTT2[i] = nLTTstat(s$newick, phylos2[[i]], distance_method = "abs")
  gam2[i] = gammaStat(phylos2[[i]])
}
nltt_values <- get_nltt_values(phylos, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)

dat = data.frame(x=c(nLTT,nLTT2),method=c(rep('method 1',length(nLTT)),rep('method 2',length(nLTT))))
ggplot(dat, aes(x=x, fill=method)) + geom_histogram(alpha=0.2, position="identity")
print(proc.time()-pp)
```

```{r}
nltt_values <- get_nltt_values(phylos2, dt = dt)
# Plot the phylognies, where the individual nLTT values are visible
qplot(t, nltt, data = nltt_values, geom = "point",
      ylim = c(0,1),
      main = "Average nLTT plot of phylogenies", size = I(0.1)
) + stat_summary(
  fun.data = "mean_cl_boot", color = "red", geom = "smooth"
)
```
