---
title: "equal likelihood trees"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dmea)
```



```{r}
st <- sim_phyl(ct=5)
plot(st$newick,root.edge = T)
axisPhylo(backward = F,root.time = st$t[1])

```



First, to warm up, we plot a simple tree
```{r echo=FALSE, include=FALSE}
library(dmea)
```

```{r}
obs = '((AB:1,AA:1):1,AC:2):1;'
#library(ape)
obs = read.tree(text=obs)
is.rooted(obs)
is.ultrametric(obs)
plot(obs,show.tip.label = T, root.edge = TRUE)
axisPhylo(backward = F,root.time = 1)
branching.times(obs)
```

Then a longer one

```{r}
# complete/incomplete trees
comp = '((AB:1,(AA:0.6,AD:0.3):0.4):1,((AE:0.6,AF:0.8):0.2,AC:1.5):0.5):1;'
comp= read.tree(text = comp)
is.rooted(comp)
is.ultrametric(comp)
plot(comp,show.tip.label = T, root.edge = TRUE)
axisPhylo(backward = F,root.time = 1)
branching.times(comp)
```


Then we drop extinxt species

```{r}
incomp = drop.fossil(comp)
write.tree(incomp)
is.rooted(incomp)
is.ultrametric(incomp)
plot(incomp,show.tip.label = T, root.edge = TRUE)
axisPhylo(backward = F,root.time = 1)
branching.times(incomp)
```

Now we would like to be able to add an extinct species

```{r}
in_dmea = phylo2p(incomp)
in_dmea$wt
in_dmea$E
in_dmea$n
ph2 <- update_tree(in_dmea,t_spe=0.5,t_ext = 1.5)
ph2o = p2phylo(ph2)
plot(ph2o,root.edge=T)
axisPhylo(backward = F,root.time = 0.5)
ltt.plot.coords(ph2o)
```

now we would like to see equivalent trees

```{r}
ph2e = p2phylo(ph2)
plot(ph2e,root.edge=T)
axisPhylo(backward = F,root.time = 0.5)
ltt.plot.coords(ph2e)
ph2e = p2phylo(ph2)
plot(ph2e,root.edge=T)
axisPhylo(backward = F,root.time = 0.5)
ltt.plot.coords(ph2e)
ph2e = p2phylo(ph2)
plot(ph2e,root.edge=T)
axisPhylo(backward = F,root.time = 0.5)
ltt.plot.coords(ph2e)
plot(ph2e,root.edge=T)
axisPhylo(backward = F,root.time = 0.5)
ltt.plot.coords(ph2e)
ph2e = p2phylo(ph2)
plot(ph2e,root.edge=T)
axisPhylo(backward = F,root.time = 0.5)
ltt.plot.coords(ph2e)
```

```{r}
s = sim_phyl(ct=6)
plot(s$newick)
p = phylo2p(s$newick)
p$E == s$E
p$wt == s$t
#p$wt
#s$t
p$n == s$n
s2 = p2phylo(p)
plot(s2)
```

voilà!, we found two things:

1. This algorithm does not label species in a general way yet, but only to models where all species has same probability to speciate.
2. We found a nice way to ilustrate 'equivalent' trees. 

```{r}
s = sim_phyl(ct=4)
plot(s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
ltt.plot.coords(s2)
nLTTstat(s2,s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
ltt.plot.coords(s2)
nLTTstat(s2,s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
ltt.plot.coords(s2)
nLTTstat(s2,s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
ltt.plot.coords(s2)
nLTTstat(s2,s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
ltt.plot.coords(s2)
nLTTstat(s2,s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
ltt.plot.coords(s2)
nLTTstat(s2,s$newick)
s2=p2phylo(phylo2p(s$newick))
plot(s2)
branching.times(s2)
nLTTstat(s2,s$newick)
plot(s$newick)
ltt.plot.coords(s2)
```

* Why the branching times are different ??
* Why on those cases the nLTT > 0 ??
