---
title: "MCEM simulations"
author: ''
date: ''
output:
  html_document: default
  pdf_document: default
---

````{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, include=FALSE,cache=FALSE}
library(dmea)
```

### Simulations


```{r, sim1}
time = proc.time()
seed = 3
s = sim_phyl(seed=3)
s2 = s$newick.extant.p
EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80),n_trees = 10, parallel = F, impsam = T,tol=0.01)
print(proc.time()-time)
```
```{r}
qplot(1:30,EM$pars[,2])
EM$pars
```

```{r, sim111}
time = proc.time()
stat=NULL
phylo1 = s$newick.extant
for(i in 1:30){
  expe = expectedLTT2(pars=EM$pars[i,])
  wt = c(expe$bt[1],diff(expe$bt))
  p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
  phylo2 = p2phylo(p)
  ltt = ltt_stat(phylo1,phylo2)
  stat[i] = ltt
}
print(proc.time()-time)
```

```{r, cache=FALSE}
 qplot(1:30,stat)
```
what if fo the expected thing with 100 trees?
```{r, ltt12}
time = proc.time()
stat=NULL
for(i in 1:30){
  expe = expectedLTT2(pars=EM$pars[i,],n_it=100)
  wt = c(expe$bt[1],diff(expe$bt))
  p = list(wt=wt,E=rep(1,(length(expe$bt)-1)),n=expe$Ex)
  phylo2 = p2phylo(p)
  ltt = ltt_stat(phylo1,phylo2)
  stat[i] = ltt
}
print(proc.time()-time)
```

```{r, cache=FALSE}
 qplot(1:30,stat)
```


wait, that was with importance sampling, now without that

```{r, sim12}
n_it=70
time = proc.time()
EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80),n_trees = 10, parallel = F, impsam = F,tol=0.01, n_it=n_it)
print(proc.time()-time)
```


```{r, cache=FALSE}
qplot(1:n_it,EM$pars[,2])
```


```{r, ltt2}
stat=NULL
time=proc.time()
for(i in 1:n_it){
  expe = expectedLTT2(pars=EM$pars[i,])
  wt = c(expe$bt[1],diff(expe$bt))
  p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
  phylo2 = p2phylo(p)
  ltt = ltt_stat(phylo1,phylo2)
  stat[i] = ltt
}
print(proc.time()-time)
```

```{r, cache=FALSE}
 qplot(1:n_it,stat)
```



I am going to do the same but with 100 trees:

nice, I want to check for another tree


```{r, sim11}
seed = 2
s = sim_phyl(seed = seed)
s2 = s$newick.extant.p
time=proc.time()
EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80),n_trees = 10, parallel = T, impsam = T,tol=0.01)
print(proc.time()-time)
```

```{r, cache=FALSE}
qplot(1:30,EM$pars[,2])
EM$pars
```


```{r, subex11}
stat=NULL
time=proc.time()
phylo1 = s$newick.extant
for(i in 1:30){
  expe = expectedLTT2(pars=EM$pars[i,])
  wt = c(expe$bt[1],diff(expe$bt))
  p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
  phylo2 = p2phylo(p)
  ltt = ltt_stat(phylo1,phylo2)
  stat[i] = ltt
}
print(proc.time()-time)
```

```{r, cache=FALSE}
 qplot(1:30,stat)
```


*** 

Complete simulations with method 1 (with the fasted way (less accuracy))

The algorithm corresponds to 

1. 




```{r, MCEM + llt}
time = proc.time()
n_sim = 100 #number of simulated trees
n_it = 30 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
for(j in 1:n_sim){
  s <- sim_phyl(seed=j)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = T, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  for(i in 1:30){
    expe = expectedLTT2(pars=EM$pars[i,])
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  Pars[j,] = EM$pars[stat==min(stat),]
  times[j] = sum(EM$times)
  qplot(1:30,stat)
}
print(proc.time()-time)
```

```{r}
summary(Pars)
qplot(Pars[,1], geom="histogram")
qplot(Pars[,2], geom="histogram")
qplot(Pars[,3], geom="histogram")
summary(time)
qplot(times, geom="histogram")
```


```{r, MCEMllt2}
time = proc.time()
n_sim = 100 #number of simulated trees
n_it = 100 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
EMs = vector(mode='list',length = n_sim) 
LTTs = vector(mode='list',length = n_sim)
for(j in 1:n_sim){
  s <- sim_phyl(seed=j)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = F, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  for(i in 1:n_it){
    expe = expectedLTT2(pars=EM$pars[i,])
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  EMs[[j]] = EM$pars
  LTTs[[j]] = stat
  Pars[j,] = EM$pars[stat==min(stat),]
  times[j] = sum(EM$times)
  print(qplot(1:n_it,stat))
}
print(proc.time()-time)
```

```{r}
summary(Pars)
qplot(Pars[,1], geom="histogram")
qplot(Pars[,2], geom="histogram")
qplot(Pars[,3], geom="histogram")
summary(time)
qplot(times, geom="histogram")
```

```{r}
for(j in 1:n_sim){
  EM=EMs[[j]]
  print(paste('´mu=0.1','seed',j))
  df = data.frame(it=1:n_it, EM=EM[,2],ltt=LTTs[[j]])
  print(EM.plot(df))
  #p = ggplot(data=df,aes(x=it,y=EM,color=ltt))+geom_point()
  #print(p)
}
```


```{r, MCEMllt3}
time = proc.time()
n_sim = 100 #number of simulated trees
n_it = 120 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
EMs = vector(mode='list',length = n_sim) 
LTTs = vector(mode='list',length = n_sim) 
for(j in 1:n_sim){
  s <- sim_phyl(seed=j,mu0 = 0.2)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = F, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  for(i in 1:n_it){
    expe = expectedLTT2(pars=EM$pars[i,])
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  EMs[[j]] = EM$pars
  LTTs[[j]] = stat
  Pars[j,] = EM$pars[stat==min(stat),]
  times[j] = sum(EM$times)
#  print(qplot(1:n_it,stat))
}
print(proc.time()-time)
```

```{r}
summary(Pars)
qplot(Pars[,1], geom="histogram")
qplot(Pars[,2], geom="histogram")
qplot(Pars[,3], geom="histogram")
summary(time)
qplot(times, geom="histogram")
```

```{r}
for(j in 1:n_sim){
  EM=EMs[[j]]
  print(paste('´mu=0.2','seed',j))
  df = data.frame(it=1:n_it, EM=EM[,2],ltt=LTTs[[j]])
  print(EM.plot(df))
}
```

```{r, MCEMllt34}
time = proc.time()
n_sim = 10 #number of simulated trees
n_it = 120 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
EMs = vector(mode='list',length = n_sim) 
LTTs = vector(mode='list',length = n_sim)
Same_dim = vector(mode='list',length=n_sim)
for(j in 1:n_sim){
  s <- sim_phyl(seed=j,mu0 = 0.2)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = F, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  same_dim = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  for(i in 1:n_it){
    expe = expectedLTT2(pars=EM$pars[i,],dim = length(s2$wt))
    same_dim[i] = expe$same_dim
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  print(proc.time()-time)
  EMs[[j]] = EM$pars
  LTTs[[j]] = stat
  Same_dim[[j]] = same_dim
  EM_sub = EM$pars[same_dim==max(same_dim),]
  stat_sub = stat[same_dim==max(same_dim)]
  choosed_stat = min(stat_sub)
  Pars[j,] = EM$pars[stat==choosed_stat,]
  times[j] = sum(EM$times)
#  print(qplot(1:n_it,stat))
}
print(proc.time()-time)
```


```{r, cache=FALSE}
Pars
summary(Pars)
```

```{r, cache=FALSE}
for(j in 1:n_sim){
  EM=EMs[[j]]
  print(paste('´mu=0.2','seed',j))
  df = data.frame(it=1:n_it, EM=EM[,2],ltt=LTTs[[j]],sd=Same_dim[[j]])
  print(EM.plot(df,dim = TRUE))
}
```


```{r, MCEMllt4}
time = proc.time()
n_sim = 100 #number of simulated trees
n_it = 80 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
EMs = vector(mode='list',length = n_sim) 
LTTs = vector(mode='list',length = n_sim) 
for(j in 1:n_sim){
  s <- sim_phyl(seed=j,mu0 = 0.4)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = F, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  for(i in 1:n_it){
    expe = expectedLTT2(pars=EM$pars[i,])
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  EMs[[j]] = EM$pars
  LTTs[[j]] = stat
  Pars[j,] = EM$pars[stat==min(stat),]
  times[j] = sum(EM$times)
#  print(qplot(1:n_it,stat))
}
print(proc.time()-time)
```

```{r}
summary(Pars)
summary(times)
```

```{r}
for(j in 1:n_sim){
  EM=EMs[[j]]
  print(paste('´mu=0.4','seed',j))
  df = data.frame(it=1:n_it, EM=EM[,2],ltt=LTTs[[j]])
  print(EM.plot(df))
  #p = ggplot(data=df,aes(x=it,y=EM,color=ltt))+geom_point()
  #print(p)
}
```


```{r, MCEMllt5}
time = proc.time()
n_sim = 10 #number of simulated trees
n_it = 150 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
EMs = vector(mode='list',length = n_sim) 
LTTs = vector(mode='list',length = n_sim) 
for(j in 1:n_sim){
  s <- sim_phyl(seed=j,mu0 = 0.4)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = F, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  for(i in 1:n_it){
    expe = expectedLTT2(pars=EM$pars[i,])
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  EMs[[j]] = EM$pars
  LTTs[[j]] = stat
  Pars[j,] = EM$pars[stat==min(stat),]
  times[j] = sum(EM$times)
#  print(qplot(1:n_it,stat))
}
print(proc.time()-time)
```

```{r}
summary(Pars)
summary(times)
```

```{r}
for(j in 1:n_sim){
  EM=EMs[[j]]
  print(paste('mu=0.4','seed',j))
  df = data.frame(it=1:n_it, EM=EM[,2],ltt=LTTs[[j]])
  print(EM.plot(df))
  #p = ggplot(data=df,aes(x=it,y=EM,color=ltt))+geom_point()
  #print(p)
}
```


```{r, MCEMllt6}
time = proc.time()
n_sim = 10 #number of simulated trees
n_it = 150 # number of EM iterations
Pars = matrix(nrow=n_sim,ncol=3)
times = vector(mode='numeric',length = n_sim) 
EMs = vector(mode='list',length = n_sim) 
LTTs = vector(mode='list',length = n_sim) 
for(j in 1:n_sim){
  s <- sim_phyl(seed=j,mu0 = 0.4)
  s2 = s$newick.extant.p
  EM = EM_phylo(wt=s2$wt, init_par = c(1.8,0.6,80), n_trees = 10, parallel = F, impsam = F, printpar = F, n_it=n_it)
  stat = vector(mode='numeric',length = n_it) 
  phylo1 = s$newick.extant
  print(paste('iteration',j))
  for(i in 1:n_it){
    expe = expectedLTT2(pars=EM$pars[i,],median=TRUE)
    wt = c(expe$bt[1],diff(expe$bt))
    p = list(wt=wt,E=rep(1,length(expe$bt)),n=expe$Ex)
    phylo2 = p2phylo(p)
    ltt = ltt_stat(phylo1,phylo2)
    stat[i] = ltt
  }
  EMs[[j]] = EM$pars
  LTTs[[j]] = stat
  Pars[j,] = EM$pars[stat==min(stat),]
  times[j] = sum(EM$times)
#  print(qplot(1:n_it,stat))
}
print(proc.time()-time)
```

```{r}
summary(Pars)
summary(times)
```

```{r}
for(j in 1:n_sim){
  EM=EMs[[j]]
  df = data.frame(n=1:n_it, EM=EM[,2],ltt=LTTs[[j]])
  p = ggplot(data=df,aes(x=n,y=EM,color=ltt))+geom_point()
  q = ggplot(data=df,aes(x=n,y=ltt,color=EM))+geom_point()
  print(paste('seed',j))
  print(p)
  print(q)
}

```

***

Simulations March 2017

Given a reconstructed phylogenetic tree

```{r}
phylo <- sim_phyl(seed=3)$newick.extant
```


The new proposed method corresponds to the following steps

1. Set initial parameters for $\lambda_0$ and $K$

```{r}
pars = c(2,80)
```


2. Given the parameters, find $\mu^*$ such that minimizes ltt_stat


```{r, sims2}
mu=0.2
# this does not work at all:
#pars = subplex(par = mu, fn = ltt_mu, phylo=phylo, prior_pars = pars)$par
#let´s visualize the ltt_mu curve
pp = proc.time()
mu = seq(0.01, 0.8, by=0.03)
stat=NULL
for(i in 1:length(mu)){
  stat[i] = ltt_mu(mu = mu[i], phylo = phylo, prior_pars = pars)
}
print(pp - proc.time())
```

and

```{r}
qplot(mu,stat)
```
```{r}
nmu = mean(mu[stat < quantile(stat,0.05)])
nmu
```

3. Once we have the new $\mu$ we can update $\lambda$ and $K$ fixing $\mu$.

HERE SHOULD BE a 2-parameter version of the EM

4. Once we have the new $\lambda$ and $K$ we go to step 2 to re-calculate $\mu$

5. We repeat this algorithm until convergence. 


***


